---
title: "Set Axis Break for ggplot2"
author: "Guangchuang Yu and Shuangbin Xu\\

        School of Basic Medical Sciences, Southern Medical University"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: github
  pdf_document:
    toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{ggbreak introduction}
  %\VignetteDepends{ggplot2}
  %\VignetteDepends{grid}
  %\VignetteDepends{aplot}
  %\VignetteDepends{ggplotify}
  %\usepackage[utf8]{inputenc}
---



```{r include=FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = TRUE)

library(ggplot2)
library(ggbreak)
library(patchwork)
library(yulab.utils)
```

## Introduction

This package was first designed to set breakpoints for truncating the plot as I need to [shrink outlier long branch of a phylogenetic tree](https://yulab-smu.top/treedata-book/faq.html#shrink-outlier-long-branch).


Axis break or a so call gap plot is useful for large datasets that are not normally distributed and contain outliers. Sometimes we can transform the data (e.g. using log-transformation if the data was log-normal distributed) to solve this problem. But this is not always granted. The data may just simply contain outliers and these outliers are meaningful. A simple gap plot can solve this issue well to present the data in detail with both normal and extreme data. 


This package provides several scale functions to break down a 'gg' plot into pieces and align them together with (gap plot) or without (wrap plot or cut plot) ignoring subplots. Our methods are fully compatible with `ggplot2`, so that users can still use the `+` operator to add geometric layers after creating a broken axis.



If you use `r CRANpkg('ggbreak')` in published research, please cite the following paper:

+ S Xu<sup>#</sup>, M Chen<sup>#</sup>, T Feng, L Zhan, L Zhou, __G Yu__<sup>\*</sup>. Use ggbreak to effectively utilize plotting space to deal with large datasets and outliers. __*Frontiers in Genetics*__. 2021, 12:774846. doi: [10.3389/fgene.2021.774846](https://www.frontiersin.org/articles/10.3389/fgene.2021.774846/full/)


## Gap plot

For creating gap plot, we only provide `scale_x_break` and `scale_y_break` functions. Currently, it is not allowed to apply both functions to set breakpoints for both x and y axes. However, multiple breakpoints on a single axis are supported.

### Feature 1: Compatible with ggplot2. 

After breaking the plot, we can still superpose geometric layers and set themes.

```{r fig.keep="last"}
library(ggplot2)
library(ggbreak) 
library(patchwork)

set.seed(2019-01-19)
d <- data.frame(x = 1:20,
   y = c(rnorm(5) + 4, rnorm(5) + 20, rnorm(5) + 5, rnorm(5) + 22)
)
 
p1 <- ggplot(d, aes(y, x)) + geom_col(orientation="y")
d2 <- data.frame(x = c(2, 18), y = c(7, 26), label = c("hello", "world"))
p2 <- p1 + scale_x_break(c(7, 17)) + 
  geom_text(aes(y, x, label=label), data=d2, hjust=1, colour = 'firebrick')  + 
  xlab(NULL) + ylab(NULL) + theme_minimal()

p1 + p2
```

### Feature 2: Multiple break-points are supported

```{r}
p2 + scale_x_break(c(18, 21))
```


### Feature 3: Zoom in or zoom out of subplots

```{r}
p1 + scale_x_break(c(7, 17), scales = 1.5) + scale_x_break(c(18, 21), scales=2)
```

### Feature 4: Support reverse scale

```{r fig.keep='last'}
g <- ggplot(d, aes(x, y)) + geom_col()
g2 <- g + scale_y_break(c(7, 17), scales = 1.5) + 
  scale_y_break(c(18, 21), scale=2) + scale_y_reverse()
g + g2
```

### Feature 5: Compatible with scale transform functions

Users can apply scale transform functions, such as `scale_x_log10` and `scale_x_sqrt`, to axis break plot.

```{r fig.keep='last', fig.width=10, fig.height=5}
p2 <- p1 + scale_x_break(c(7, 17)) 
p3 <- p1 + scale_x_break(c(7, 17)) + scale_x_log10()
p2 + p3
```

### Feature 6: Compatible with `coord_flip`

```{r message=FALSE}
g + coord_flip() + scale_y_break(c(7, 18))
```


### Feature 7: Compatible with `facet_grid` and `facet_wrap`

```{r message=FALSE}
set.seed(2019-01-19)
d <- data.frame(
  x = 1:20,
  y = c(rnorm(5) + 4, rnorm(5) + 20, rnorm(5) + 5, rnorm(5) + 22),
  group = c(rep("A", 10), rep("B", 10)),
  face=c(rep("C", 5), rep("D", 5), rep("E", 5), rep("F", 5))
)

p <- ggplot(d, aes(x=x, y=y)) +
     geom_col(orientation="x") +
     scale_y_reverse() +
     facet_wrap(group~.,
                scales="free_y",
                strip.position="right",
                nrow=2
                ) +
     coord_flip()
pg <- p +
  scale_y_break(c(7, 17), scales="free") +
  scale_y_break(c(19, 21), scales="free")
print(pg)
```



### Feature 8: Compatible with legends

```{r message=FALSE}
pg <- pg + aes(fill=group) + theme(legend.position = "bottom")
print(pg)
```

### Feature 9: Supports all plot labels

```{r message=FALSE}
pg + labs(title="test title", subtitle="test subtitle", tag="A tag", caption="A caption") +
     theme_bw() +
     theme(
           legend.position = "bottom",
           strip.placement = "outside",
           axis.title.x=element_text(size=10),
           plot.title = element_text(size = 22),
           plot.subtitle = element_text(size = 16),
           plot.tag = element_text(size = 10),
           plot.title.position = "plot",
           plot.tag.position = "topright",
           plot.caption = element_text(face="bold.italic"),

     )
```

### Feature 10: Allows setting tick labels for subplots

```{r message=FALSE, fig.width=10, fig.height=6}
require(ggplot2)
library(ggbreak)
set.seed(2019-01-19)
d <- data.frame(
  x = 1:20,
  y =  c(rnorm(5) + 4, rnorm(5) + 20, rnorm(5) + 5, rnorm(5) + 22),
  group = c(rep("A", 10), rep("B", 10))
)

p <- ggplot(d, aes(x=x, y=y)) +
     scale_y_reverse() +
     scale_x_reverse() +
     geom_col(aes(fill=group)) +
     scale_fill_manual(values=c("#00AED7", "#009E73")) +
     facet_wrap(
         group~.,
         scales="free_y",
         strip.position="right",
         nrow=2
     ) +
     coord_flip()                                                                                                                                                                                                  

p +
     scale_y_break(c(7, 10), scales=0.5, ticklabels=c(10, 11.5, 13)) +
     scale_y_break(c(13, 17), scales=0.5, ticklabels=c(17, 18, 19)) +
     scale_y_break(c(19,21), scales=1, ticklabels=c(21, 22, 23))
```

### Feature 11: Compatible with dual axis

```{r fig.width=10, fig.height=5}
p <- ggplot(mpg, aes(displ, hwy)) +
     geom_point() +
     scale_y_continuous(
       "mpg (US)",
       sec.axis = sec_axis(~ . * 1.20, name = "mpg (UK)")
     ) +
      theme(
        axis.title.y.left = element_text(color="deepskyblue"),
        axis.title.y.right = element_text(color = "orange")
      )
p1 <- p + scale_y_break(breaks = c(20, 30))
p2 <- p + scale_x_break(breaks = c(3, 4))
p1 + p2
```

### Feature 12: Compatible with patchwork

```{r message=FALSE, fig.width=8, fig.height=5, fig.keep="last"}
library(patchwork)

set.seed(2019-01-19)
d <- data.frame(
               x = 1:20,
               y = c(rnorm(5) + 4, rnorm(5) + 20, rnorm(5) + 5, rnorm(5) + 22)
)

p <- ggplot(d, aes(x, y)) + geom_col()
x <- p+scale_y_break(c(7, 17 ))

x + p
```

## Wrap plot

The `scale_wrap()` function wraps a 'gg' plot over multiple rows to make plots with long x-axes easier to read.

```{r fig.width=6, fig.height=8}
p <- ggplot(economics, aes(x=date, y = unemploy, colour = uempmed)) +
  geom_line()

p + scale_wrap(n=4)
```


Both categorical and numerical variables are supported.

```{r fig.width=7, fig.height=6}
ggplot(mpg, aes(class, hwy)) + 
  geom_boxplot() +
      scale_wrap(n = 2)
```

## Cut plot

The `scale_x_cut` or `scale_y_cut` cuts a 'gg' plot to several slices with the ability to specify which subplots to zoom in or zoom out.

```{r}
library(ggplot2)
library(ggbreak)
set.seed(2019-01-19)
d <- data.frame(
     x = 1:20,
     y = c(rnorm(5) + 4, rnorm(5) + 20, rnorm(5) + 5, rnorm(5) + 22)
 )
p <- ggplot(d, aes(x, y)) + geom_col()
p + scale_y_cut(breaks=c(7, 18), which=c(1, 3), scales=c(3, 0.5))
```

## Adjust the amount of space between subplots

The `space` parameter in `scale_x_break()`, `scale_y_break()`, `scale_x_cut()` and `scale_y_cut()` allows user to control the space between subplots.


```{r}
p + scale_y_cut(breaks=c(7, 18), which=c(1, 3), scales=c(3, 0.5), space=.5)
```

## Place legend at any position


```{r legend-subview, fig.keep="last"}
## original plot
p1 <- ggplot(mpg, aes(displ, hwy, color=factor(cyl))) + geom_point()

## ggbreak plot without legend
p2 <- p1 + scale_x_break(c(3, 4)) +
    theme(legend.position="none") 

## extract legend from original plot
leg = cowplot::get_legend(p1)

## redraw the figure
p3 <- ggplotify::as.ggplot(print(p2))

## place the legend 
p3 + ggimage::geom_subview(x=.9, y=.8, subview=leg)
```



## Note

The features we introduced for `scale_x_break` and `scale_y_break` also work for `scale_wrap`, `scale_x_cut` and `scale_y_cut`.

## FAQ

1. Incompatible with functions that arrange multiple plots

You can use `aplot::plot_list()` to arrange `ggbreak` objects with other `ggplot` objects. For other functions, such as `cowplot::plot_grid()` and `gridExtra::grid.arrange()`, you need to explictly call `print()` to `ggbreak` object, see also <https://github.com/YuLab-SMU/ggbreak/issues/36>.

Using `print()` is a secret magic to make `ggbreak` compatible with other packages, including [export](https://github.com/YuLab-SMU/ggbreak/issues/37). 

